Think deep for reef resilience

### 2025-05-30 paper/paper_draft.qmd
Add paper qmd edits and possible carbonate factored cover analyses

---
title: Future climate resilience of the Great Barrier Reef
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations: 
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
    custom:
    - kind: float
      reference-prefix: SFigure
      key: sfig
---

@@

ecs_width = fig_sizes["ecs_width"]
timeseries_width = fig_sizes["timeseries_width"]
violin_width = fig_sizes["violin_width"]
```

@@

Climate change is impacting species survival, persistence and ecosystem functioning globally, including coral. Coral reefs are vulnerable to increasing magnitude and variation in environmental stressors such as ocean temperature, acidification and nutrient loading ([@foster2021]; [@gattuso2014]). Mass bleaching and mortality events linked to climate change can cause reef-building coral population declines, possibly leading to long-term and large-scale coral-macroalgae regime shifts (Arif et al. 2022; Wild et al. 2011). Climate impacts on coral reefs are, and will continue to, decrease the ability of coral reefs to provide ecological functions such as benthic habitat and nutrient cycling (Foster & Attrill 2021; Williams & Graham 2019) as well as ecosystem services such as food, protection, tourism and Indigenous cultural value (Eddy et al. 2021; Stoeckl et al. 2021). We aim to assess the impacts of possible climate futures and identify the characteristics of resilient reefs using the latest generation of larval connectivity research with scenario-based population modelling for the Great Barrier Reef ecosystem.

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of thousands of individual reefs. They are interconnected through coral and fish larval dispersal patterns mediated by hydrodynamic, spatial, and ecological processes (Grimaldi et al. 2022). Severe declines in coral cover are projected due to compounding impacts such as ocean warming and acidification, intensification of Tropical Cyclones, and Crown of Thorns Starfish outbreaks (De’ath et al. 2012; Fox-Kemper et al. 2021; Seneviratne et al. 2021; Sweatman et al. 2011; Wolff et al. 2018). Ominously, the 2023-24 Summer mass bleaching event (Cantin et al. 2024) may be signalling a shift towards this reef future.

Environmental abiotic factors influence how reefs will experience climate change pressures, including the spatio-temporal patterns of thermal stress itself, and indirect factors that influence the stress experienced by corals. Environmental factors such as tropical cyclones and wind driven mixing (Lin et al. 2023; McClanahan et al. 2005); currents (Grimaldi et al. 2022) and depth (Álvarez-Noriega et al. 2018; Baird et al. 2018; Smith et al. 2014) can decrease the magnitude of thermal stress corals experience. On the other hand, high light intensity and nutrient levels in the benthic zone can increase coral bleaching (DeCarlo et al. 2020; Higuchi et al. 2015). These environmental factors influence the resilience of reefs to climate change pressures in combination with related biological factors such as larval connectivity and thermal tolerance.

Connectivity in systems such as the GBR is an important contributor to system resilience, as strong connections are required to repopulate locations after disturbance, and drive adaptation to climate pressures (Angeler et al. 2023; Elmhirst et al. 2009; Hock et al. 2017). When populations that are strong larval suppliers to external reefs are impacted by disturbances, increased adult coral mortality will cause decreased outgoing larvae in the following years. Weaker larval connections then slow the recovery of the externally connected locations (Figueiredo et al. 2022; Hock et al. 2017; Munday et al. 2009). Populations with greater strength and number of outgoing than incoming larval connections - herein referred to as larval sources - have lower resilience to pressures than populations with greater incoming than outgoing larval flows - referred to as larval sinks - as larval source populations lack external sources of larvae that are important in contributing to reef recovery and recruitment of juveniles after disturbances to existing populations (Ani et al. 2024; Hock et al. 2017). Larval source populations are an important component of overall reef resilience as they can contribute larvae to recovering populations at other locations (Hock et al. 2017; Thomas et al. 2015). Connectivity will be altered by future climate change impacts such as population fragmentation, increasing larval mortality, and changes to coastal hydrodynamics (Figueiredo et al. 2022; Munday et al. 2009), and influences the biology of surrounding reef populations.

Many additional biological factors including coral size (Brandt 2009), taxa (Álvarez-Noriega et al. 2018; McClanahan et al. 2005), and acclimation and inherited thermal tolerance (Humanes et al. 2022; Lachs et al. 2024), influence the outcome of high coral heat stress. It has been shown that corals have the adaptive capacity to develop thermal tolerance levels capable of delaying climate impacts under low emissions scenarios (Bairos-Novak et al. 2021; Humanes et al. 2022).

Incorporating the latest advancements in whole ecosystem larval connectivity modelling and research on the mechanisms behind coral bleaching and thermal adaptation responses into coral population modelling allows us to investigate how these abiotic and biotic factors interact and will influence coral climate change resilience into the future. Improvements in coral reef climate response modelling, including sampling from a range of possible climate futures, allow for fine-scale timeseries analyses of uncertain climate change and ecological scenarios (Bode et al. 2017; Chollett et al. 2022; Donner et al. 2018; Klein et al. 2024; Uusitalo et al. 2015).

@@

## Environmental modelling context

The spatial extent of GBR coral reefs used in this investigation includes 3806 reef areas, as used in ReefMod-GBR (Bozec et al. 2022). Annually and spatially averaged Degree Heating Week (DHW) data is used to represent future thermal stress at each reef location. Five Global Climate Models (GCMs) (EC-Earth3-Veg, ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and GDFL-CM4) from the Sixth Phase of the Coupled Model Intercomparison Project (CMIP6) were selected for this study. Statistically downscaled 1km resolution Sea Surface Temperature (SST) data for the Great Barrier Reef were obtained from the HighResCoralStress database (Dixon et al. 2023, 2022). Maximum monthly mean DHW levels were calculated from the SST data following NOAA methodology (Liu et al. 2017; Skirving et al. 2020). These GCMs were selected from available GCMs downscaled by Dixon et al. (2023) based on evaluation by Grose et al. (2023), which found high correlation between observed temperature, rainfall, El Niño Southern Oscillation and Indian Ocean Dipole patterns in the Australian terrestrial region, and the corresponding mechanisms in these GCMs. Equilibrium climate sensitivity (ECS) is defined, by the IPCC, as the long term warming caused by a doubling of carbon dioxide above its pre-industrial concentration (Forster et al. 2021). The four GCMs span a range of ECS values within the IPCC assessed likely range, and GCM development families, allowing for a range of possible climate futures to be explored, capturing the uncertainty in future responses to CO2 increases (@fig-ecs; Grose et al. 2023; Shepherd et al. 2018). The following analyses are performed using each GCM separately. SSP 2-4.5 is used as a plausible emission pathway under current emission trajectories, with similar atmospheric CO2 levels remaining until 2050 (Hausfather & Peters 2020; IPCC 2023).

@@

CoralBlox is a coral population growth and mortality model that simulates groups of corals growing in diameter and competing for limited space over time (CoralBlox citation). This model incorporates linear extension (growth) and background mortality rates that are dependent on the size and functional group of a coral colony. Environmental disturbances and reproduction are not directly considered within CoralBlox. Instead, these factors are considered by making changes to the distribution of colonies within CoralBlox timesteps using the ADRIA (Adaptive Dynamic Reef Intervention Algorithms) platform (ADRIA citation). Within ADRIA, we incorporate coral life history and environmental events in the following order: coral colony growth; coral reproduction; larval connectivity and settlement; and coral bleaching due to heat stress.

CoralBlox uses user-defined size classes that influence size-dependent growth and mortality rates, through which the coral colonies progress as they grow in diameter. Within ADRIA, we define the distributions of these parameters, and the size classes based on ecological monitoring data from the [EcoRRAP program](https://gbrrestoration.org/program/ecorrap/).

Coral fecundity rates are dependent on colony size and functional group following the results of Hall & Hughes (1996). Coral fertilization rates are not considered within this model, however larval survival and viability are incorporated in connectivity modelling (see Connectivity modelling). Viable larvae settle at a reef location and are incorporated into the model as the smallest size class, but are not directly impacted by modelled bleaching mortality as it is assumed the high background mortality of juveniles includes thermal stress.

@@

We track the thermal tolerance of each functional group and size class as a truncated normal distribution that is updated in each timestep following growth and recruitment. This distribution determines that proportion of the population on a reef that is susceptible to bleaching (i.e. has a thermal threshold below the experienced DHW value). Starting distributions for thermal tolerance levels are informed by Bairos-Novak et al. (2021) and unpublished data from Hughes et al. (2018). Tropical cyclones and wave impacts to reefs are not included in this investigation. We scale coral bleaching impacts by a coefficient determined by the median depth of each reef location, following Baird et al. (2018), reducing the proportion of coral colonies on a reef impacted by high heat stress as depth increases. Median depths for each reef location have been extracted from satellite derived bathymetry data for the Great Barrier Reef (GBRMPA 2020). Due to varying degrees of alignment between bathymetry data and the defined outlines for GBR reefs, reefs with less than 5% coverage of valid bathymetry data have been removed from this investigation. Furthermore, reefs that are not allocated to a GBRMPA management area or bioregion have been removed from the investigation, as well as reefs with relative coral cover values greater than five.

@@

\*\* need information on GBR1 connectivity used \*\*

@@

Timeseries analysis is performed by first grouping reefs into three different scales, GBRMPA reefal bioregion, GBRMPA management area and GBR-wide, to investigate the consistency of cluster identification. To identify groups of reefs at each analysis scale that are resilient, intermediate and non-resilient we clustered reef relative coral cover timeseries into three clusters at each scale. To quantify the behaviour of timeseries and identify timeseries that behave similarly we calculate the complexity-invariant distance (CID) between each timeseries. The complexity-invariant distance refers to the Euclidean distance between two timeseries multiplied (corrected) by the ratio of the complexities of the timeseries (Batista et al. 2014). These values create a distance matrix of n \* n dimensions (where n is the number of timeseries in the group) that is clustered using a k-medoids clustering algorithm (Steinmann et al. 2020).

@@

Reefs in each cluster are compared to identify characteristics that differ between low, medium and high coral cover clusters. These characteristics include reef median depth, mean timeseries DHW, mean timeseries DHW – tolerance level, Pearson’s correlation between total coral cover and inversed DHW timeseries, proportion of coral habitable area occupied by corals in the first timestep, ratio of source-to-sink larval connectivity and total strength of larval connections between reefs. To analyse the DHW - tolerance levels of reefs, heritable DHW – tolerance values were tracked over timeseries in a timestep \* taxa \* reef \* scenario matrix. The mean timeseries was calculated over the first four scenarios for each GCM (four from 4096 scenarios per GCM) due to computational restraints, and then used to calculate an overall mean DHW – tolerance value for each reef under each GCM.

@@

Within bioregions timeseries clustering analysis was able to identify three clusters of relative cover timeseries. Across many bioregions timeseries clustering identified a ‘low’ cluster containing reefs that suffer larger declines and generally have lower levels of relative coral cover than the other clusters, a ‘medium’ cluster with intermediate levels of relative coral cover, and a ‘high’ cluster with cover recovery that persists to 2050 in many bioregions (@fig-access-esm1-5-bioregion-timeseries).

![Reef relative coral cover timeseries clustered within each bioregion. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each bioregion. Solid lines represent the cluster median timeseries and bands represent the 2.5 to 97.5 percentile confidence intervals. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/bioregion_cover_timeseries.png){#fig-access-esm1-5-bioregion-timeseries width=`{julia} timeseries_width`}

In many bioregions, median reef depth decreases as cluster relative coral cover decreases, with reefs within the ‘high’ cluster commonly having the greatest median reef depths within the bioregion (@fig-access-esm1-5-bioregion-depth).

@@

Correlation between reef total coral cover and reef DHW timeseries decreases as cluster relative coral cover decreases, with reefs in the ‘high’ cluster commonly having the greatest total cover – DHW timeseries correlation levels (@fig-access-esm1-5-bioregion-dhwcor).

![Total coral cover – DHW Pearson’s correlation levels for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each bioregion. Solid lines represent the median cover – DHW correlation level for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/dhw_cover_cor_bioregion_violin.png){#fig-access-esm1-5-bioregion-dhwcor width=`{julia} violin_width`}

Conversely, there are no consistent patterns across bioregions in reef relative cover and mean timeseries DHW levels, with clusters having similar mean DHW levels in many bioregions (@fig-access-esm1-5-bioregion-dhwmean).

@@

In some bioregions (4 for ACCESS-ESM1-5) reefs in the ‘low’ cluster have substantially lower mean thermal tolerance levels compared to reefs in other clusters. However, many bioregions do not show strong differences in thermal tolerance levels (@fig-access-esm1-5-bioregion-dhwtol).

![Mean timeseries DHW tolerance value for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each bioregion. Solid lines represent the median DHW tolerance value for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/dhw_tolerance_bioregion_violin.png){#fig-access-esm1-5-bioregion-dhwtol width=`{julia} violin_width`}

In many bioregions reefs with high relative coral cover levels (‘high’ cluster) have the lowest initial proportions of habitable area occupied by corals, while reefs in the ‘low’ cluster often have the highest initial proportions of coral habitable area occupied by corals (@fig-access-esm1-5-bioregion-initialproportion).

![Initial proportion of coral habitable area occupied by corals for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each bioregion. Solid lines represent the median proportion of habitable area occupied for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/initial_proportion_bioregion_violin.png){#fig-access-esm1-5-bioregion-initialproportion width=`{julia} violin_width`}

Across most bioregions there are not consistent differences between clusters in connectivity metrics such as source-to-sink connectivity ratio (@fig-access-esm1-5-bioregion-logsosi) and total connectivity strength (@sfig-access-esm1-5-bioregion-totalconnectivity).

![Log source – to – sink connectivity ratios for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each bioregion. Solid lines within each violin represent the median log source – to – sink ratio for each cluster and grey points represent the ratio value for each reef. Vertical grey lines represent a log source – to – sink ratio of 0 indicating equal outgoing and incoming connectivity levels. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/so_to_si_bioregion_violin.png){#fig-access-esm1-5-bioregion-logsosi width=`{julia} violin_width`}

## Management area scale clustering
Within the four management areas, timeseries clustering analysis identified three clusters of reefs using relative coral cover timeseries. These clusters form groups of reefs with ‘low, medium and high’ relative coral cover, categorised by their median coral cover between 2030 and 2040. The ‘high’ cover cluster of reefs contains reefs that have continuing recovery of corals throughout the future climate scenario. The medium and low clusters contain many reefs that decrease to half their initial coral cover levels and largely do not display the levels of recurring recovery that reefs in high clusters experience (@fig-access-esm1-5-manarea-timeseries).

@@

# Supplementary Material
## ACCESS-ESM1-5

>>>>

---
title: Think deep for reef resilience
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations: 
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
    attributes:
        corresponding: true
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier Reef (GBR) experiencing severe declines in coral cover from thermal stress events. Understanding which reefs exhibit resilience to climate impacts is critical for conservation planning and intervention strategies. A scenario-based population modelling approach was adopted with larval connectivity dynamics, thermal adaptation, and environmental factors to assess coral reef resilience across GBR locations under future climate projections. Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities, three distinct reef trajectories are identified: resilient reefs maintaining coral cover through 2050, intermediate reefs with moderate declines, and vulnerable reefs experiencing severe degradation. Contrary to expectations, larval connectivity patterns did not substantially differentiate resilient from vulnerable reefs. Instead, reef depth emerged as the primary predictor of resilience, with deeper reefs benefiting from thermal refugia that reduced bleaching impacts. Initial coral cover also influenced trajectories, with reefs having lower initial cover showing greater relative resilience. Our findings suggest that depth provides critical protection against thermal stress, contrasting current restoration practices that primarily target easily accessible shallow sites. These results have important implications for reef management, indicating that depth refugia should be prioritized for further consideration in conservation planning alongside connectivity considerations to maximize coral survival under continuing climate change.
keywords: 
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
    custom:
    - kind: float
      reference-prefix: SFigure
      key: sfig
---

@@

map_width = fig_sizes["map_width"]
ecs_width = fig_sizes["ecs_width"]
timeseries_width = fig_sizes["timeseries_width"]
violin_width = fig_sizes["violin_width"]
```

@@

Climate change is impacting species survival, persistence and ecosystem functions globally, including coral. Coral reefs are vulnerable to increasing magnitude and variation in environmental stressors such as ocean temperature, acidification and nutrient loading ([@foster2021]; [@gattuso2014]). Mass bleaching and mortality events linked to climate change can cause reef-building coral population declines, possibly leading to long-term and large-scale coral-macroalgae regime shifts ([@arif2022]; [@wild2011]). Climate impacts on coral reefs are, and will continue to, decrease the ability of coral reefs to provide ecological functions such as benthic habitat and nutrient cycling ([@forster2021]; [@williams2019]) as well as ecosystem services such as food, protection, tourism and Indigenous cultural value ([@eddy2021]; [@stoeckl2021]). We aim to assess the impacts of possible climate futures and identify the characteristics of resilient reefs using the latest generation of larval connectivity research with scenario-based population modelling for the Great Barrier Reef ecosystem.

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of thousands of individual reefs. They are interconnected through coral and fish larval dispersal patterns mediated by hydrodynamic, spatial, and ecological processes ([@grimaldi2022]). Severe declines in coral cover are projected due to compounding impacts such as ocean warming and acidification, intensification of Tropical Cyclones, and Crown of Thorns Starfish outbreaks ([@death2012]; [@fox-kemper2021]; [@seneviratne2021]; [@sweatman2011]; [@wolff2018]). Ominously, the 2023-24 Summer mass bleaching event ([@cantin2024]) may be signalling a shift towards this reef future, with some level of bleaching recorded on 79% of surveyed reefs.

Environmental abiotic factors influence how reefs will experience climate change pressures, including the spatio-temporal patterns of thermal stress itself, and indirect factors that influence the stress experienced by corals. Environmental factors such as tropical cyclones and wind driven mixing ([@lin2023]; [@mcclanahan2005]); currents ([@grimaldi2022]) and depth ([@álvarez-noriega2018]; [@baird2018]; [@smith2014]) can decrease the magnitude of thermal stress corals experience. On the other hand, high light intensity and nutrient levels in the benthic zone can increase coral bleaching ([@decarlo2020]; [@higuchi2015]). These environmental factors influence the resilience of reefs to climate change pressures in combination with related biological factors such as larval connectivity and thermal tolerance.

Connectivity in systems such as the GBR is an important contributor to system resilience, as strong connections are required to repopulate locations after disturbance, and drive adaptation to climate pressures ([@angeler2023]; [@elmhirst2009]; [@hock2017]). When populations that are strong larval suppliers to external reefs are impacted by disturbances, increased adult coral mortality will cause decreased outgoing larvae in the following years. Weaker larval connections then slow the recovery of the externally connected locations ([@figueiredo2022]; [@hock2017]; [@munday2009]). Populations with greater strength and number of outgoing than incoming larval connections - herein referred to as larval sources - have lower resilience to pressures than populations with greater incoming than outgoing larval flows - referred to as larval sinks - as larval source populations lack external sources of larvae that are important in contributing to reef recovery and recruitment of juveniles after disturbances to existing populations ([@ani2024]; [@hock2017]). Larval source populations are an important component of overall reef resilience as they can contribute larvae to recovering populations at other locations ([@hock2017]; [@thomas2015]). Connectivity will be altered by future climate change impacts such as population fragmentation, increasing larval mortality, and changes to coastal hydrodynamics ([@figueiredo2022]; [@munday2009]), and influences the biology of surrounding reef populations.

Many additional biological factors including coral size ([@brandt2009]), taxa ([@álvarez-noriega2018]; [@mcclanahan2005]), and acclimation and inherited thermal tolerance ([@humanes2022]; [@lachs2024]), influence the outcome of high coral heat stress. It has been shown that corals have the adaptive capacity to develop thermal tolerance levels capable of delaying the onset of widespread bleaching under low emissions scenarios ([@bairos-novak2021]; [@humanes2022]).

Incorporating the latest advancements in whole ecosystem larval connectivity modelling and coral thermal response and adaptation research into coral population modelling allows us to investigate how these abiotic and biotic factors interact and will influence coral climate change resilience into the future. Improvements in coral reef climate response modelling, including sampling from a range of possible climate futures, allow for fine-scale timeseries analyses of uncertain climate change and ecological scenarios ([@bode2017]; [@chollett2022]; [@donner2018]; [@klein2024]; [@uusitalo2015]).

@@

## Environmental modelling context
![Map of the Great Barrier Reef region, including GBRMPA management areas (polygons) and all considered reefs (points). The green polygon represents the Far Northern management area, the yellow polygon represents the Cairns/Cooktown management area, the pink polygon represents the Townsville/Whitsunday management area and the blue polygon represents the Mackay/Capricorn management area. The centroid of each reef is represented as a point for visual clarity, and is coloured according to its GBRMPA reefal bioregion. An annotation represents the approximate location of the Australian Institute of Marine Science Townsville site at Cape Cleveland.](/../outputs/ADRIA_results/HighResCoralStress/figs/region_map.png){#fig-region-map width=`{julia} map_width`}

The spatial extent of GBR coral reefs used in this investigation includes 3806 reef areas. It follows the reef definitions as adopted in ReefMod-GBR ([@bozec2022]) (@fig-region-map). Annually and spatially averaged Degree Heating Week (DHW) data is used to represent future thermal stress at each reef location. Five Global Climate Models (GCMs) (EC-Earth3-Veg, ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and GDFL-CM4) from the Sixth Phase of the Coupled Model Intercomparison Project (CMIP6) were selected for this study. Statistically downscaled 1km resolution Sea Surface Temperature (SST) data for the Great Barrier Reef were obtained from the HighResCoralStress database ([@dixon2022][@dixon]). Maximum monthly mean DHW levels were calculated from the SST data following NOAA methodology ([@liu2017]; [@skirving2020]). These GCMs were selected from available GCMs downscaled by Dixon et al. [-@dixon] based on evaluation by Grose et al. [-@grose2023], which found high correlation between observed temperature, rainfall, El Niño Southern Oscillation and Indian Ocean Dipole patterns in the Australian terrestrial region, and the corresponding mechanisms in these GCMs. Equilibrium climate sensitivity (ECS) is defined, by the IPCC, as the long term warming caused by a doubling of carbon dioxide above its pre-industrial concentration ([@forster2021]). The four GCMs span a range of ECS values within the IPCC assessed likely range, and GCM development families, allowing for a range of possible climate futures to be explored, capturing the uncertainty in future responses to CO2 increases (@fig-ecs; [@grose2023]; [@shepherd2018]). The following analyses are performed using each GCM separately. SSP 2-4.5 is used as a plausible emission pathway under current emission trajectories, with similar atmospheric CO2 levels remaining until 2050 ([@hausfather2020]; IPCC [-@climate2023]).

@@

CoralBlox is a coral population growth and mortality model that simulates groups of corals growing in diameter and competing for limited space over time ([@ribeirodealmeida2024]). This model incorporates linear extension (growth) and background mortality rates that are dependent on the size and functional group of a coral colony. Environmental disturbances and reproduction are not directly considered within CoralBlox. Instead, these factors are considered by making changes to the distribution of colonies within CoralBlox timesteps using the ADRIA (Adaptive Dynamic Reef Intervention Algorithms) platform ([@iwanaga2023]). Within ADRIA, we incorporate coral life history and environmental events in the following order: coral colony growth; coral reproduction; larval connectivity and settlement; and coral bleaching due to heat stress. As ADRIA and CoralBlox are coupled together to represent the system, we refer to both as ADRIA-CoralBlox.

Coral fecundity rates are dependent on colony size and functional group following the results of Hall & Hughes [-@hall1996]. Coral fertilization rates are not considered within this model, however larval survival and viability are incorporated in connectivity modelling (see Connectivity modelling). Viable larvae settle at a reef location and are incorporated into the model as the smallest size class but are not directly impacted by modelled bleaching mortality as it is assumed the high background mortality of juveniles includes thermal stress.

@@

We track the thermal tolerance of each functional group and size class as a truncated normal distribution that is updated in each timestep following growth and recruitment. This distribution determines that proportion of the population on a reef that is susceptible to bleaching (i.e. has a thermal threshold below the experienced DHW value). Starting distributions for thermal tolerance levels are informed by Bairos-Novak et al. [-@bairos-novak2021] and unpublished data from Hughes et al. [-@hughes2018].

Tropical cyclones and wave impacts to reefs are not included in this investigation. We scale coral bleaching impacts by a coefficient determined by the median depth of each reef location, following Baird et al. [@baird2018], reducing the proportion of coral colonies on a reef impacted by high heat stress as depth increases. Median depths for each reef location have been extracted from satellite derived bathymetry data for the Great Barrier Reef (GBRMPA [-@gbr10gb2020]) (see discussion for more details). Due to varying degrees of alignment between bathymetry data and the defined outlines for GBR reefs, reefs with less than 5% coverage of valid bathymetry data have been removed from this investigation. Furthermore, reefs that are not allocated to a GBRMPA management area or bioregion have been removed from the investigation, as well as reefs with relative coral cover values greater than five.

@@

** need information on GBR1 connectivity used **

@@

Timeseries analysis is performed by first grouping reefs into three different scales, GBRMPA reefal bioregion, GBRMPA management area and GBR-wide, to investigate the consistency of cluster identification. To identify groups of reefs at each analysis scale that are resilient, intermediate and non-resilient we clustered reef relative coral cover timeseries into three clusters at each scale. To quantify the behaviour of timeseries and identify timeseries that behave similarly we calculate the complexity-invariant distance (CID) between each timeseries. The complexity-invariant distance refers to the Euclidean distance between two timeseries multiplied (corrected) by the ratio of the complexities of the timeseries ([@batista2014]). These values create a distance matrix of n * n dimensions (where n is the number of timeseries in the group) that is clustered using a k-medoids clustering algorithm ([@steinmann2020]).

@@

Reefs in each cluster are compared to identify characteristics that differ between low, medium and high coral cover clusters. These characteristics include reef median depth, mean timeseries DHW, mean timeseries DHW – tolerance level, Pearson’s correlation between total coral cover and inversed DHW timeseries, proportion of coral habitable area occupied by corals in the first timestep, ratio of source-to-sink larval connectivity and total strength of larval connections between reefs. To analyse the DHW - tolerance levels of reefs, heritable DHW – tolerance values were tracked over timeseries in a timestep * taxa * reef * scenario matrix. The mean timeseries was calculated over the first four scenarios for each GCM (four from 4096 scenarios per GCM) due to computational restraints, and then used to calculate an overall mean DHW – tolerance value for each reef under each GCM.

@@

Within bioregions timeseries clustering analysis was able to identify three clusters of relative cover timeseries. Across many bioregions timeseries clustering identified a ‘low’ cluster containing reefs that suffer larger declines and generally have lower levels of relative coral cover than the other clusters, a ‘medium’ cluster with intermediate levels of relative coral cover, and a ‘high’ cluster with cover recovery that persists to 2050 in many bioregions (@sfig-access-esm1-5-bioregion-timeseries; @sfig-access-cm2-bioregion-timeseries; @sfig-ec-earth3-veg-bioregion-timeseries; @sfig-gfdl-cm4-bioregion-timeseries; @sfig-noresm2-mm-bioregion-timeseries).

In many bioregions, median reef depth decreases as cluster relative coral cover decreases, with reefs within the ‘high’ cluster commonly having the greatest median reef depths within the bioregion (@fig-access-esm1-5-bioregion-depth; @sfig-access-cm2-bioregion-depth; @sfig-ec-earth3-veg-bioregion-depth; @sfig-gfdl-cm4-bioregion-depth; @sfig-noresm2-mm-bioregion-depth).

@@

Correlation between reef total coral cover and reef DHW timeseries decreases as cluster relative coral cover decreases, with reefs in the ‘high’ cluster commonly having the greatest total cover – DHW timeseries correlation levels (@sfig-access-esm1-5-bioregion-dhwcor; @sfig-access-cm2-bioregion-dhwcor; @sfig-ec-earth3-veg-bioregion-dhwcor; @sfig-gfdl-cm4-bioregion-dhwcor; @sfig-noresm2-mm-bioregion-dhwcor).

Conversely, there are no consistent patterns across bioregions in reef relative cover and mean timeseries DHW levels, with clusters having similar mean DHW levels in many bioregions (@fig-access-esm1-5-bioregion-dhwmean; @sfig-access-cm2-bioregion-dhwmean; @sfig-ec-earth3-veg-bioregion-dhwmean; @sfig-gfdl-cm4-bioregion-dhwmean; @sfig-noresm2-mm-bioregion-dhwmean).

@@

In some bioregions (4 for ACCESS-ESM1-5) reefs in the ‘low’ cluster have substantially lower mean thermal tolerance levels compared to reefs in other clusters. However, many bioregions do not show strong differences in thermal tolerance levels (@sfig-access-esm1-5-bioregion-dhwtol; @sfig-access-cm2-bioregion-dhwtol; @sfig-ec-earth3-veg-bioregion-dhwtol; @sfig-gfdl-cm4-bioregion-dhwtol; @sfig-noresm2-mm-bioregion-dhwtol).

In many bioregions reefs with high relative coral cover levels (‘high’ cluster) have the lowest initial proportions of habitable area occupied by corals, while reefs in the ‘low’ cluster often have the highest initial proportions of coral habitable area occupied by corals (@sfig-access-esm1-5-bioregion-initialproportion; @sfig-access-cm2-bioregion-initialproportion; @sfig-ec-earth3-veg-bioregion-initialproportion; @sfig-gfdl-cm4-bioregion-initialproportion; @sfig-noresm2-mm-bioregion-initialproportion).

Across most bioregions there are not consistent differences between clusters in connectivity metrics such as source-to-sink connectivity ratio (@sfig-access-esm1-5-bioregion-logsosi; @sfig-access-cm2-bioregion-logsosi; @sfig-ec-earth3-veg-bioregion-logsosi; @sfig-gfdl-cm4-bioregion-logsosi; @sfig-noresm2-mm-bioregion-logsosi) and total connectivity strength (@sfig-access-esm1-5-bioregion-totalconnectivity; @sfig-access-cm2-bioregion-totalconnectivity; @sfig-ec-earth3-veg-bioregion-totalconnectivity; @sfig-gfdl-cm4-bioregion-totalconnectivity; @sfig-noresm2-mm-bioregion-totalconnectivity).

## Management area scale clustering

Within the four management areas, timeseries clustering analysis identified three clusters of reefs using relative coral cover timeseries. These clusters form groups of reefs with ‘low, medium and high’ relative coral cover, categorised by their median coral cover between 2030 and 2040. The ‘high’ cover cluster of reefs contains reefs that have continuing recovery of corals throughout the future climate scenario. The medium and low clusters contain many reefs that decrease to half their initial coral cover levels and largely do not display the levels of recurring recovery that reefs in high clusters experience (@fig-access-esm1-5-manarea-timeseries; @sfig-access-cm2-manarea-timeseries; @sfig-ec-earth3-veg-manarea-timeseries; @sfig-gfdl-cm4-manarea-timeseries; @sfig-noresm2-mm-manarea-timeseries).

@@

# Supplementary Material

## Model parameterisation
CoralBlox uses user-defined size classes that influence size-dependent growth and mortality rates, through which the coral colonies progress as they grow in diameter. Within ADRIA, we define the distributions of these parameters, and the size classes based on ecological monitoring data from the Ecological Intelligence for Reef Restoration Program (EcoRRAP) ([@australian-institute-of-marine-science2025]; [@ribeirodealmeida2024]).
